How is experience transformed into memory?

### 2018-09-02 paper/main.tex
draft of discussion complete
working on general cleanup

The topics we found were heavily character-focused (e.g., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes that were primarily Sherlock-focused; primarily John-focused (John is Sherlock's close confidant and assistant); or that involved Sherlock and John interacting (Fig.~\topics).  Several of the topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences (if the distinctions between those overlapping topics were meaningful).  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) topics tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistance} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B).  Following \cite{BaldEtal17}, we used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow).  Part of our model fitting procedure required selecting an appropriate number of ``events'' to segment the timeseries into.  We used an optimization procedure to identify the number of events that maximized within-event stability while also minimizing across-event correlations (see \textit{Methods} for additional details).  To create a stable ``summary'' of the video, we computed the average topic vector within each event (Fig.~\ref{fig:model}C).

@@

One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the movie).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlaped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in ][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectory).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiencs; ][]{KestEtal12, GilbMarl17}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.

Our work has broad implications for how we characterize and assess memory in real-world settings such as the classroom or physician's office.  For example, the most commonly used classroom evaluation tools involve computing the proportion of correctly answered exam questions.  Our work indicates that this approach is only loosly related to what educators might really want to measure: how well did the students understand the key ideas presented in the course?  One could apply the computational framework we developed to construct topic trajectories for the video and participants' recalls to build explicit content models of the course material and exam questions.  This approach would provide a more nuanced and specific view into which aspects of the material students had learned well (or poorly).  Similarly, memory evaluations that incorporate explicit content models might provide more direct measures of the qualitiy of patients' memories.

% JRM THOUGHTS ON DISCUSSION:
% - Nature says "1-2 short paragraphs" -- needs to be much shorter.
% - I think we can get rid of anything we mention in the intro or results
% - Things not easily presented elsewhere are the background on the brain results and applications to education metrics.  I think we can focus on those two parts, add an "overview" and "conclusion" (potentially), and cut the rest.

>>>>

The topics we found were heavily character-focused (e.g., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes that were primarily Sherlock-focused; primarily John-focused (John is Sherlock's close confidant and assistant); or that involved Sherlock and John interacting (Fig.~\topics).  Several of the topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences (if the distinctions between those overlapping topics were meaningful; also see Fig.~\featureimportance).  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) topics tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistance} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B).  Following \cite{BaldEtal17}, we used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow).  Part of our model fitting procedure required selecting an appropriate number of ``events'' to segment the timeseries into.  We used an optimization procedure to identify the number of events that maximized within-event stability while also minimizing across-event correlations (see \textit{Methods} for additional details).  To create a stable ``summary'' of the video, we computed the average topic vector within each event (Fig.~\ref{fig:model}C).

@@

One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the movie).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlaped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectory).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiencs; ][]{KestEtal12, GilbMarl17}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.

Our work has broad implications for how we characterize and assess memory in real-world settings such as the classroom or physician's office.  For example, the most commonly used classroom evaluation tools involve computing the proportion of correctly answered exam questions.  Our work indicates that this approach is only loosly related to what educators might really want to measure: how well did the students understand the key ideas presented in the course?  One could apply the computational framework we developed to construct topic trajectories for the video and participants' recalls to build explicit content models of the course material and exam questions.  This approach would provide a more nuanced and specific view into which aspects of the material students had learned well (or poorly).  In clinical settings, memory measures that incorporate such explicit content models might also provide more direct evaluations patients' memories.
